Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations

### 2017-08-09 TexSubmission/answer_referees.tex

\vspace{.3cm}

\item {\em I quite like the choice of $Q(\theta;\theta)$ in the examples. It would be very interesting to
have an implementation of the actual EM algorithm for parameter estimation in the
examples provided, using GRand Paris on the E step.}

\vspace{.3cm}

The numerical section now contains a complete implementation of a generalized EM algorithm for the SINE model which highlights clearly the performance of the proposed PaRIS algorithm.  
\end{enumerate}

### 2017-08-09 TexSubmission/ParisEM.tex
Added GEM application

\subsection*{The SINE model} 
The performance of the GRand PaRIS algorithm are first highlighted using the SINE model, where $(X_t)_{t\geq 0}$ is supposed to be the solution to: 
\begin{equation}
\rmd X_t = \sin \left(X_t-\theta\right)\rmd t + \rmd W_t,~~X_0=x_0\eqsp. \label{eq:Lamp:SINE}
\end{equation}
This simple model has no explicit transition density, however GPE estimators may be computed by simulating Brownian bridges.
The process solution to \eqref{eq:Lamp:SINE} is observed regularly at times $t_0=0,\ldots,t_{100}=50$ through the observation process $(Y_k)_{0\leq k \leq 100}$:
\begin{equation}
Y_k = X_k + \varepsilon_k\label{eq:obs:SINE}\eqsp,
\end{equation}
where the $(\varepsilon)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0,1)$.
In the example displayed on Figure \ref{fig:res:SINE:obs}, we set $\theta=0$.
In that case, the function $\rho_{\Delta_k}$ defined in \eqref{eq:rho:func} can be upper bounded either on $(x,y)$ or only on $y$, the GRand PaRIS algorithm has therefore a linear complexity.

@@

>>>>

\subsection*{The SINE model} 
The performance of the GRand PaRIS algorithm are first highlighted using the SINE model, where $(X_t)_{t\geq 0}$ is supposed to be the solution to: 
\begin{equation}
\rmd X_t = \sin \left(X_t-\theta\right)\rmd t + \rmd W_t,~~X_0=x_0\eqsp. \label{eq:Lamp:SINE}
\end{equation}
This simple model has no explicit transition density, however GPE estimators may be computed by simulating Brownian bridges.
The process solution to \eqref{eq:Lamp:SINE} is observed regularly at times $t_0=0,\ldots,t_{100}=50$ through the observation process $(Y_k)_{0\leq k \leq 100}$:
\begin{equation}
Y_k = X_k + \varepsilon_k\label{eq:obs:SINE}\eqsp,
\end{equation}
where the $(\varepsilon_k)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0,1)$.
In the example displayed on Figure \ref{fig:res:SINE:obs}, we set $\theta=0$.
In that case, the function $\rho_{\Delta_k}$ defined in \eqref{eq:rho:func} can be upper bounded either on $(x,y)$ or only on $y$, the GRand PaRIS algorithm has therefore a linear complexity.

@@

The performance of our algorithm are also assessed in the case where $\theta$ and the variance $\sigma^2$ of $\varepsilon_1$ are unknown and estimated using a generalized EM algorithm: $n=200$ observations are sampled with $\theta = 0$ and $\sigma^2 = 0.5$ and the Grand PaRIS algorithm is based on $N=200$ particles, $\widetilde{N} = \textcolor{red}{???}$ backward samples and $M = \textcolor{red}{???}$ replicates of the transition density estimator to approximate the importance weights. 
As their is no closed form solution to compute the M step of the EM algorithm and propose new parameter estimates, at each iteration $50$ new parameter estimates are sampled \textcolor{red}{comment ?} close to the current estimate and the intermediate quantity is approximated using our algorithm for each candidate. Then, parameter values leading to the greatest value of the approximate intermediate quantities are selected as the new parameter estimates.
Figures~\ref{fig:SINE:theta} and~\ref{fig:SINE:sigma} illustrate the performance of the estimation for several initializations of $\theta$ (resp. $\sigma^2$) uniformly chosen in  $(-\pi,\pi)$ (resp. in $(0,6)$) after only a few iterations of the EM procedure.

\begin{figure}[p]
\centering
\includegraphics[scale=0.5]{figure_theta}
\caption{{\em SINE model - EM}. Estimation  of $\theta$.}
\label{fig:SINE:theta}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[scale=0.5]{figure_sigma2}
\caption{{\em SINE model - EM}. Estimation of $\sigma^2$.}
\label{fig:SINE:sigma}
\end{figure}
